Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-22 doc/ApTranscriptome_manuscript.Rmd
updates to ms

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: pnas1.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.4°C for *A. picea*; Supplemental Material) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*; Table S1) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2014-10-08.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these 9,052 transcripts, 80% (`r prettyNum(length(quadratic.lms), big.mark = ",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.

@@

Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Table 1, Figure 1). *Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, post-hoc $\chi^2$ = ...stats...), whereas *A. carolinensis* downregulated more transcripts under these conditions (*Intermediate*, post-hoc $\chi^2$ = ...stats...). Consistent with the *enhanced defense* hypothesis, *A. picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A. carolinensis* (Table 1, $\chi^2$ = ...). In contrast, at high temperatures, *A. carolinensis* matched the predictions of the *tolerance hypothesis*, inducing expression of ~14% fewer transcripts than *A. picea* (Table 1).

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes. **Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (...stat...), whereas there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

The molecular processes in the set of transcripts that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), transcript regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by examining whether the suite of transcripts induced in response to either temperature extreme showed a higher constitutive level of expression, here defined as expression at 19.5°C, in the species better adapted to these conditions. Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A. picea* for transcripts that were responsive to high temperature in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`). However, the reverse pattern of higher constitutive expression in *A. picea* was not observed for low temperature responsive transcripts in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`).

Instead, the overall pattern is that thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than having constitutively elevated expression as predicted, these genes were uniformly not expressed (...mean TPM...) in *A. carolinensis*.

...GO terms **High** and **Bimodal** transcripts in *A. picea* that were **Not Responsive** in *A. carolinensis* ...

@@

@@

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperature extremes: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 2A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Supplemental...). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* might need to induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the short growing season that would necessitate foraging under broader ranges of temperature (...stuble data...).

@@

@@

![Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences.](../results/matched_observations_Fig1.png)

![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*.](../results/crit_temp_regulation_Fig2.png)

![Scatterplots of (A) constitutive expression in *A. picea* against degree of induction for upregulated transcripts in *A. carolinensis* and (B) constitutive expression in *A. carolinensis* against degree of induction for upregulated transcripts in *A. picea.*. For each species, the transcripts are split into expression category based on the ... The top row is for *High* transcripts, middle is for *Intermediate* transcripts and bottom row is for *Low* transcripts.](../results/constitutive_vs_upreg_Fig3.png)

>>>>

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: pnas1.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.4°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_slim_2014-10-22.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these 9,052 transcripts, 80% (...) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.

@@

The two species differed in the number of transcripts in each response category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`). *Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A. carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`). Consistent with the *enhanced defense* hypothesis, *A. picea* induced 1,146 (30%) more transcripts in response to low temperatures than *A. carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`). However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, at high temperatures *A. carolinensis* induced expression of ~14% fewer transcripts than *A. picea* (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* < `r max(chi5$p.value, 0.001)`).

```{r echo=FALSE, results='hide'}
# results of Stuart-Maxwell test
mh.test
```

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Table 1, Figure S1) indicating that these differences are not simply due to neutral processes. **Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (Fig. 1A), whereas there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1B). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

![Deviations from expected numbers of transcripts in matched observation data for transcript expression type data. The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit). Shown are **Low**, **High** and **Bimodal** expression types for (A) *A. picea** and (B) *A. carolinensis* against all expression types in the opposite species. The dashed boxes indicate the diagonals (e.g. High - High) which are positive](../results/matched_observations_Fig1.png)

The molecular processes in the set of transcripts that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism. In addition, there were GO terms related to repair (e.g. histone modification), transcript regulation (e.g. positive regulation of cellular process) and cellular organization (e.g. transport, organelle organization). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

We tested the *genetic assimilation hypothesis*, a shift from inducible to constitutive expression of stress-response genes, by examining whether the suite of transcripts induced in response to either temperature extreme showed higher constitutive expression, here defined as expression at 19.5°C, in the species better adapted to these conditions. This would be evident as a positive relationship between the ratio of inducibility between species against the opposite ratio of constitutive expression between species. However, for both high temperature activated transcripts in *A. picea* (Fig. 2A) and low temperature activated transcripts in *A. carolinensis* (Fig. 2B) there was no evidence for greater constitutive expression for transcripts with lower inducibility. In fact, for both comparisons there was the opposite relationship as predicted by the *genetic assimilation hypothesis*, with greater inducibility for transcripts with lower constitutive expression in the opposite species.

![Scatterplots of (A) log ratio of inducibility in *A. picea* over inducibility in *A. carolinensis* against log ration of constitutive expression in *A. carolinensis* over constitutive expression in *A. picae* for transcripts with **High** expression in *A. picea*, and (B) log ratio of inducibility in *A. carolinensis* over inducibility in *A. picea* against log ration of constitutive expression in *A. picea* over constitutive expression in *A. carolinensis*. In both cases, a positive relationship would be Lines are ordinary least squares regressions on the data.](../results/genetic_assimilation_Fig2.png)

In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A. carolinensis*.

@@

![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*.](../results/crit_temp_regulation_Fig3.png)

@@

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperature extremes: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 2A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* might need to induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the short growing season that would necessitate foraging under a broader range of temperatures.

@@

One possible explanation for the differences between species in their thermal reactionomes is neutral drift of gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. While we cannot rule this possibility out, two lines of evidence make us think that it is unlikely. First, if gene expression differed by drift between species, we would not expect to see any striking differences among expression type classes as we do here (Fig. 1, Fig. S1). Second, even if these differences were due to drift, it would be unlikely that there would be transcriptome-wide differences in the critical temperatures of gene activation such as we see here (Fig. 2).

@@
